\section{Limitations and Roadmap}
\label{sec:limitations}

\subsection{Evaluation limitations}

The benchmark has only five scenarios and 33 assertions. They are deterministic, public, and visible to the implementer. Passing them cannot estimate a field attack rate, robustness to paraphrase, multilingual injection, multimodal content, long-horizon drift, or adaptive adversaries. The flagship demonstration is one carefully chosen vertical slice. Its value is inspectable causality, not breadth.

The benchmark and project share a maintainer, and the ``independent'' verifiers share a repository and author. An external team should rerun the artifacts, inspect adapter semantics, add hidden scenarios, and implement at least one verifier from the written specification.

The public comparison mixes native memory agents, managed services, direct APIs, store-level harnesses, and conversation-session persistence. A 33/33 harness may include explicit adapter policy not present in every default product path. The paper therefore avoids broad competitive claims.

\subsection{Implementation limitations}

\paragraph{No universal action gateway.} Memory MCP exists, but guarded actions are not yet exposed as a mandatory MCP interception layer. An agent with another filesystem or network tool can bypass the reference engine.

\paragraph{Host-attested, not cryptographically authenticated, source and authority.} The engine trusts source type and CLI authority flags supplied by the host. Subject IDs are storage scopes, not authenticated tenants.

\paragraph{Shared-key approval.} HMAC is a practical local primitive but lacks signer non-repudiation and makes key distribution/rotation operationally sensitive. Approver labels are claims.

\paragraph{One narrow adapter.} The filesystem vertical slice supports two operations. Email, browser, database, cloud, messaging, payment, and code-execution adapters need provider-specific preconditions, idempotency, observation, and compensation.

\paragraph{Adapter-coupled verification.} The same adapter family prepares, executes, and verifies. Separate observer credentials and implementations would reduce correlated failure.

\paragraph{Deterministic extraction and lexical retrieval.} Generic patterns and FTS5 make behavior reproducible but limit paraphrase coverage. Model-backed extraction remains future work and must preserve quarantine semantics.

\paragraph{Storage and deployment maturity.} SQLite is the only storage backend. Payload encryption, remote multi-tenant service hardening, KMS integration, role-based administration, secure checkpoint publication, and operational observability remain deployment work.

\paragraph{No formal proof or machine-checked model.} Properties in this report are code-level arguments under stated assumptions. The state machine and schemas have not been verified in TLA+, Alloy, Coq, Lean, or a model checker.

\subsection{Roadmap}

The highest-leverage next steps are:

\begin{enumerate}
  \item \textbf{Complete mediation:} add an authenticated action gateway for MCP/HTTP hosts and remove direct mutation capabilities from agent-facing processes.
  \item \textbf{Asymmetric approval:} define a signer interface for KMS/HSM/passkey-backed signatures, key identifiers, rotation, revocation, and role policy.
  \item \textbf{Independent observation:} support read-only verifier principals and provider audit-log correlation distinct from executor credentials.
  \item \textbf{Provider adapters:} implement conditional/idempotent adapters for email, calendars, Git, cloud storage, databases, browsers, and messaging.
  \item \textbf{Formal protocol publication:} version and publish \worldpatch{}, approval, receipt, and verifier schemas as a permissively licensed protocol package.
  \item \textbf{Broader evaluation:} add guarded-action tracks to \benchmark{}, hidden/adaptive scenarios, multiple models/hosts, crash injection, concurrency, fuzzing, and repeated performance experiments.
  \item \textbf{Policy-preserving graph memory:} broaden deterministic relation extraction and multi-hop evaluation behind the same admission, provenance, deletion, and retrieval-evidence boundaries.
  \item \textbf{External witnessing:} integrate transparency logs, WORM storage, or trusted timestamping for production checkpoint anchors.
\end{enumerate}

The design target is not ``a better prompt.'' It is a deployable control path in which the model never possesses all authority required to approve, execute, and rewrite the evidence for its own effects.
