\section{Implementation Evidence and Scale Probe}
\label{sec:evaluation}

This paper evaluates implementation consistency, the recall-forensics task of \cref{sec:forensics}, and one bounded-recall scenario. It does not claim state-of-the-art answer quality. The committed scale evidence in \code{paper/data/governed-memory-probe.json} comes from release v0.3.0. The forensics evidence in \code{paper/data/recall-forensics.json} comes from the content-addressed post-release snapshot identified by its embedded 42-file digest manifest.

\subsection{Questions and method}

The probe asks three narrow questions: (Q1) does direct scoring remain capped after an older relevant record is surrounded by 10,000 newer records; (Q2) can the graph recover a two-hop target that direct top-10 recall omits; and (Q3) what local latency overhead does graph recall add in this scenario?

The harness creates an in-memory SQLite store, remembers ``my boss is Sarah'' and ``Sarah's preferred airport is SEA,'' then adds 10,000 synthetic slot records. It issues ``What does my boss use for flights?'' with a result limit of 10 and a direct-candidate cap of 200. Three warm-up pairs precede 25 measured direct/graph pairs. Latencies exclude ingestion and graph construction. The p95 reported by the script is an order statistic, not an interpolated estimate.

The host was an Apple M2 with 16\,GiB RAM, macOS Darwin 25.5.0, and Python 3.12.2 running as an x86-64 process. The source commit was \code{69ebedea93ea95965867af5483dd9f63425b8e75}.

\begin{table}[t]
\centering
\caption{10,002-record local probe. Times are milliseconds.}
\label{tab:probe}
\small
\begin{tabularx}{\columnwidth}{@{}Yrr@{}}
\toprule
Measure & Direct & Graph \\
\midrule
Median recall & 17.139 & 24.799 \\
p95 order statistic & 17.417 & 25.394 \\
Target rank in top 10 & absent & 2 \\
Candidate cap & 200 & 200 + bounded spread \\
Logged candidate summaries & \multicolumn{2}{r}{50} \\
Graph edges visited & -- & 2 \\
\bottomrule
\end{tabularx}
\end{table}

Ingestion of all 10,002 records took 5,621.290\,ms in this in-memory run. Graph recall added 7.660\,ms to the median (about 45\% relative overhead) and recovered the target at rank 2. Direct recall omitted it because FTS reached the ``boss'' edge but the target answer record shared no query term and was older than the 200-record recent fallback window. The result supports the intended multi-hop mechanism for this construction; one query cannot establish general retrieval quality.

\subsection{Invariant tests}

On the evaluated source tree, \code{python3 -m pytest -q} completed 155 tests (the final timing is reported in the evidence provenance file). The suite includes focused checks for retrieval-envelope binding, score inputs, result limits above the 50-row diagnostic window, and graph rebuild determinism, in addition to admission policy, quarantine and promotion, correction and deletion, FTS and graph lifecycle, path recall, reversible merges, archive digests, incremental audit verification, approvals, precondition changes, uncertain outcomes, compensation, root confinement, service roles, and macOS sealing behavior. Passing unit and integration tests demonstrates conformance to the encoded cases, not absence of security defects.

\subsection{What supports scale, and what is unproven}

Three implementation choices prevent obvious linear work in the hot recall path: direct candidate scoring is capped; graph traversal has explicit seed/frontier/depth bounds; and only inactive derived edges are eligible for cold subject/year partitioning. FTS and graph row maps permit targeted updates and deletions. Incremental audit verification avoids rehashing a validated prefix during routine checks.

These properties make larger stores technically plausible without changing canonical semantics. They do not establish a supported maximum. This release has no committed million-record benchmark, no sustained concurrent-write test, no long-running archive churn study, and no representative multilingual or paraphrase corpus. SQLite file size, FTS selectivity, cache pressure, audit volume, backup duration, and maintenance time remain empirical constraints. Claims beyond the 10,002-record probe are design projections.

\subsection{Reproduction}

From the repository root:

\begin{lstlisting}[language=bash]
python3 -m pytest -q
python3 paper/scripts/run_governed_memory_probe.py \
  --records 10000 --iterations 25 --candidate-cap 200
python3 paper/scripts/run_recall_forensics_eval.py \
  --filler 1000 --seed 20260716
cd paper && make aetnamem-governed-memory.pdf
\end{lstlisting}

Timing results should be expected to vary across hosts and runs. Rank, cap, path, and visited-edge assertions are the more reproducible outputs. The recall-forensics counts in \cref{tab:forensics} are deterministic for the file-digest manifest, fixed workload, and seed; the reconstruction time is a single-host observation.
