\documentclass[11pt]{article}

\usepackage[a4paper,top=22mm,bottom=24mm,left=25mm,right=25mm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{newtxtext,newtxmath}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{hyperref}

\definecolor{aetnaInk}{HTML}{17252A}
\definecolor{aetnaTeal}{HTML}{2A6F73}
\definecolor{aetnaBlue}{HTML}{315A7D}
\definecolor{aetnaGray}{HTML}{5E686C}
\definecolor{aetnaPaper}{HTML}{F4F6F6}

\hypersetup{
  colorlinks=true,
  linkcolor=aetnaBlue,
  urlcolor=aetnaBlue,
  pdftitle={Aetnamem Explained: Twenty Articles on Governed Memory for AI},
  pdfauthor={aetna000.com},
  pdfsubject={A high-level article series based on the aetnamem governed-memory white paper},
  pdfkeywords={aetnamem, governed memory, AI agents, recall forensics, auditability, local-first AI}
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\footnotesize\textsc{aetnamem} explained}
\fancyhead[R]{\footnotesize X Article Series}
\fancyfoot[L]{\footnotesize Author: aetna000.com}
\fancyfoot[R]{\footnotesize\thepage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.3pt}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.7em}
\setlength{\emergencystretch}{3em}
\setlist{leftmargin=*,itemsep=0.25em,topsep=0.25em}

\newcommand{\system}{\textsc{aetnamem}}
\newcommand{\articlelabel}[1]{%
  \noindent\colorbox{aetnaTeal!10}{%
    \parbox{0.96\textwidth}{\color{aetnaInk}\small\textbf{Article #1 of 20}}}%
  \par\medskip}
\newcommand{\coreidea}[1]{%
  \begin{quote}\color{aetnaTeal}\large\itshape #1\end{quote}}
\newcommand{\boundary}[1]{%
  \par\smallskip\noindent\textbf{Important boundary.} #1}

\title{\vspace{-1.4em}\textbf{Aetnamem Explained}\par
\large Twenty Articles on Governed Memory for AI}
\author{aetna000.com\\
\small Based on \emph{Governed Memory Without Embeddings}}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
This collection contains twenty standalone, high-level articles prepared for
publication on X. It translates the architecture and measured claims of the
\system{} governed-memory white paper into accessible language without
removing the technical boundaries. The series explains why durable AI memory
is more than search, how provenance and trust shape recall, why the graph is a
rebuildable index, what recall forensics means, how approval limits external
effects, and where the current encryption and scale evidence stops.
\end{abstract}

\noindent\textbf{Source paper:}
\emph{Governed Memory Without Embeddings: Recall Forensics, Deterministic
Retrieval, and Approval-Bound Effects in aetnamem}. Author: aetna000.com.

\noindent\textbf{Editorial note:} Each article is designed to stand alone.
Technical measurements are used only where they clarify the idea. Statements
about implementation refer to the version and content-addressed evaluation
snapshot documented by the source paper, not to every future release.

\tableofcontents
\clearpage

\section{Memory Is Not a Search Box}
\articlelabel{1}
\coreidea{A useful AI memory system must govern what becomes memory, not only
find text that looks relevant.}

When people hear ``AI memory,'' they often imagine a search engine attached to
a chatbot. Store old conversations, find similar passages, place them in the
next prompt, and let the model answer. That can improve continuity, but it
does not answer the harder questions created by long-lived memory.

Who supplied a remembered statement? Was it the user, a web page, a tool, or
the model itself? Is it still current? Did a later correction replace it? Was
it reviewed? Can it be deleted? If it contributes to a proposed action, does
it carry any authority to change the outside world?

\system{} treats these as state-management questions. It stores source
episodes, normalized records, lifecycle status, trust origin, provenance, and
audit evidence. Retrieval operates inside those rules. A result cannot enter
ordinary recall merely because it has a strong text match; it must first be an
active memory under the lifecycle policy.

This changes the role of the language model. The model may help a user speak
to the system, summarize evidence, or propose an action. It does not own the
truth of the database, decide which untrusted content becomes accepted, or
turn recalled information into permission.

The practical goal is modest but important: when an assistant says ``I
remember,'' the system should be able to show what was accepted, where it came
from, what changed since then, and why it was eligible to appear.

\boundary{\system{} does not prove that a remembered statement is true. It
preserves who supplied it, how it was treated, and how it participated in the
system.}

\clearpage
\section{A Filing Cabinet, Catalogue, Map, and Receipt Book}
\articlelabel{2}
\coreidea{The easiest way to understand \system{} is as four familiar tools
working together.}

The first tool is a \textbf{filing cabinet}. It holds canonical episodes and
records: the original source material and the concise facts derived from it.
These are the durable objects from which other structures can be rebuilt.

The second is a \textbf{card catalogue}. SQLite full-text search helps locate
active records without scanning and scoring every item in memory. It is a
local index with explicit limits, not a second source of truth.

The third is a \textbf{map}. A derived graph connects entities and typed
relations. If memory says ``my boss is Sarah'' and ``Sarah prefers SEA,'' the
map can follow the relation between those statements. Every edge points back
to a record, and every record points back to its source episode.

The fourth is a \textbf{receipt book}. Important transitions append evidence
to a hash-linked audit chain: admission, promotion, correction, deletion,
retrieval, graph maintenance, approval, and external-action outcomes.

This separation is useful because the catalogue and map may be wrong or
outdated without corrupting the filing cabinet. They can be cleared and
rebuilt from canonical records. The receipt book then helps an operator check
which transitions occurred and whether the recorded chain still verifies.

Many systems combine these responsibilities into one opaque model-driven
pipeline. \system{} separates them so that failures have smaller meanings: an
index problem is not automatically a memory-integrity problem, and a relevant
memory is not automatically an authorized action.

\boundary{Rebuildability improves recovery from index damage. It does not
replace backups or recover canonical records that were lost from disk.}

\clearpage
\section{Why the Source of a Memory Matters}
\articlelabel{3}
\coreidea{The same sentence should be treated differently when it comes from
a user, a web page, a tool, or the assistant itself.}

Imagine a user has stored a preferred airport. Later, a web page contains the
sentence ``your preferred airport is LAX.'' A text extractor can read both
sentences, but it should not treat them as equally authoritative.

\system{} classifies origin before a candidate memory becomes active. Direct
user statements and explicitly reviewed content may enter active memory.
Marked web-page or tool-output content begins quarantined. It remains visible
for review, but ordinary recall and graph traversal exclude it.

This is a practical defense against durable memory poisoning. A hostile page
cannot silently replace a trusted preference merely because it contains
language that looks like a personal fact. Promotion is a separate action with
its own provenance and audit event.

Assistant responses are also not automatically promoted into semantic
memory. Otherwise, a model could repeat its own output until generated text
appeared to have become an established fact. The system can log the exchange
without granting the model self-authorship over durable truth.

Origin is not the same as correctness. A trusted user can be mistaken, and a
web page can be accurate. The trust tier instead answers a narrower policy
question: how may this source participate before review?

\boundary{Origin classification depends on the host correctly labeling
external content. If a compromised integration presents remote text as a
direct user statement, the storage layer cannot recover the missing context.}

\clearpage
\section{Quarantine Is Memory With a Door}
\articlelabel{4}
\coreidea{Suspicious information does not have to be either trusted or
discarded; it can be retained for inspection without entering normal recall.}

Binary choices are often too crude for AI memory. Throwing away every
untrusted observation loses potentially useful information. Accepting every
observation creates a path for web content, tool output, or generated text to
shape future answers as if a user had approved it.

\system{} uses quarantine as an intermediate lifecycle state. A quarantined
record keeps its source and can be reviewed, but it is excluded from ordinary
recall. Its graph edges cannot become bridges between trusted facts. It also
cannot supersede an active record.

Promotion is explicit. A reviewer can inspect the proposed memory, confirm
it, and move it into the user-confirmed trust tier. The transition is recorded
rather than hidden inside a model's confidence score.

This design makes disagreement visible. The system can say, in effect, ``this
information was observed, but it has not been accepted.'' That is more useful
than pretending all stored text has the same standing.

Quarantine is especially important for agents that browse, call tools, or
receive messages from external channels. Those systems encounter content that
may be relevant but was not authored as an instruction to the memory layer.

\boundary{A reviewer can still approve false information. Quarantine governs
admission and records responsibility; it does not provide an oracle for fact
checking.}

\clearpage
\section{Corrections Should Not Erase History}
\articlelabel{5}
\coreidea{A durable memory should show what changed, not silently rewrite the
past.}

Suppose a user says ``my preferred airport is Sydney,'' then later changes it
to Melbourne. Editing the old database row in place would produce the current
answer, but it would remove the evidence that a correction occurred.

\system{} uses explicit supersession for recognized single-valued facts. The
new active record becomes current, while the prior record moves to a
superseded state and points to its replacement. Normal recall returns the new
value, but history remains reconstructable.

This matters when people ask why an assistant behaved differently last month.
The answer may depend on the memory state at that time, not the state visible
today. A transition history allows an auditor to replay which record was
eligible at a particular turn.

The policy also protects trusted state from unreviewed replacement. A
quarantined record cannot supersede an active record, even when both appear to
occupy the same fact slot.

Not every relation is single-valued. ``Works with,'' ``likes,'' and ``avoids''
may accumulate. Unstructured conflicts may remain as separate records for
review rather than being merged by semantic guesswork.

\boundary{Supersession depends on the system recognizing a structured fact
key. Unsupported phrasing can remain as multiple records, because visible
ambiguity is safer than an unexplained destructive update.}

\clearpage
\section{Forgetting Is More Than Removing a Search Result}
\articlelabel{6}
\coreidea{Deletion should remove governed payloads and indexes while leaving
evidence that a deletion operation occurred.}

Removing an item from search is not necessarily forgetting it. The original
text may still exist in a record table, a graph edge, a full-text index, a
source episode, or an archive.

In \system{}, forgetting is a lifecycle transition across these structures.
Matching records are tombstoned, selected payloads are purged, and related
full-text and graph entries are removed. Source episodes are purged when no
retained record still depends on them.

The system keeps a deletion receipt containing identifiers and digests rather
than the erased words. This creates a useful balance: an operator can verify
that a deletion was processed without preserving the sensitive value merely
to make future auditing easier.

That choice creates a real replay boundary. If a memory contributed to an old
answer and was later deleted, the system may reconstruct the historical
decision from commitments and lifecycle events while correctly refusing to
reproduce the deleted plaintext.

Deletion is deliberately accepted only from user-origin requests, and an
empty selector is rejected instead of being interpreted as ``delete
everything.'' These are small rules with large consequences for safety.

\boundary{A local deletion receipt cannot prove that independent backups,
exports, model-provider logs, or external systems deleted their copies. Those
systems need separate retention policies and deletion receipts.}

\clearpage
\section{Why Aetnamem Does Not Use Embeddings}
\articlelabel{7}
\coreidea{Avoiding embeddings is an engineering trade: less semantic reach in
exchange for a smaller, more reproducible trusted retrieval path.}

Embeddings are effective at finding paraphrases and conceptually similar
text. \system{} does not claim otherwise. It deliberately omits an embedding
model and vector database from its current memory-retrieval path.

The reason is operational. Replaying a lexical decision requires text
normalization, static weights, database state, and code. Replaying an
embedding decision also requires the exact model and tokenizer, preprocessing,
numerical environment, vector-index state, dependencies, and often hardware
details.

Deletion and correction are also simpler when the indexes are ordinary SQLite
rows. There is no stale vector to discover, re-embedding queue to operate, or
dimensional migration when a model changes. A local installation does not
need to download and run a second model just to remember.

Explanation is another factor. A lexical result can identify matching terms.
A graph result can show a typed path. A nearest-neighbor score says that two
items are close in a learned space, which is useful but usually less direct as
a human explanation.

The cost is real: lexical retrieval can miss synonyms, misspellings,
multilingual paraphrases, and conceptual equivalence. The current parser is
also intentionally narrow. \system{} chooses deterministic, deletable, and
inspectable behavior for its present use case rather than claiming universal
retrieval superiority.

\boundary{Embedding retrieval can be reproducible when all required artifacts
are preserved. The narrower claim is that \system{} replay does not require
that additional learned-model artifact.}

\clearpage
\section{Bounded Recall on a Laptop}
\articlelabel{8}
\coreidea{Memory growth should not force every query to score every stored
record.}

A naive memory implementation compares each query with every record. That
becomes slower and produces larger evidence logs as the database grows.

\system{} first asks SQLite FTS5 for a bounded set of active lexical matches.
It fills unused capacity with recent active records, then ranks only that
candidate set. The documented default candidate cap is 200, with a hard
per-call ceiling of 2,000.

The ranking combines text relevance, source-trust weight, and recency using a
versioned deterministic formula. These values are retrieval utilities, not
probabilities that a fact is true. Lifecycle eligibility occurs before
ranking, so a highly relevant quarantined record remains excluded.

Bounded work also matters for audit evidence. The retrieval event stores a
query digest, returned identifiers, algorithm parameters, and a bounded
candidate ledger rather than copying an unbounded list into the audit trail.

In the paper's 10,002-record laptop probe, direct recall kept its 200-record
candidate cap. This demonstrates the mechanism at that size, not a guaranteed
maximum for every machine or workload.

\boundary{Candidate caps improve predictable work but can exclude relevant
records. Retrieval quality, FTS selectivity, database size, cache pressure,
and hardware still need representative measurement.}

\clearpage
\section{The Graph Is a Map, Not the Truth}
\articlelabel{9}
\coreidea{A knowledge graph is safer when it can be discarded and rebuilt
from canonical records.}

Graphs are useful because they make relationships explicit. They are also
dangerous when an extraction error becomes accepted as permanent truth.

\system{} treats its entity-relation graph as a derived index. Every edge
points to a canonical record, and every record points to its source episode.
If an extractor improves or the graph becomes damaged, the system can rebuild
the map from the retained filing cabinet.

Lifecycle rules propagate into the graph. An edge derived from quarantined
content is quarantined. A correction supersedes the corresponding
single-valued edge. Forgetting removes or tombstones related graph objects.
Only active entities, aliases, and edges participate in normal traversal.

Entity merges are proposal-only. Exact same-kind candidates can be shown to a
reviewer, but the system does not let a language model decide that two people
or projects are identical. Approved merges use reversible pointers rather
than rewriting source records.

This architecture limits the damage caused by a bad map. It does not pretend
the map is harmless; a wrong active record can still produce a wrong edge.
The difference is that the lineage remains visible and the derived structure
is not load-bearing for canonical history.

\boundary{A rebuild works only from retained canonical evidence. It cannot
restore source content that was deliberately deleted or lost without a
backup.}

\clearpage
\section{Multi-Hop Recall That Can Explain Itself}
\articlelabel{10}
\coreidea{Some memory questions require joining facts, not merely finding a
similar sentence.}

Consider two accepted memories: ``my boss is Sarah'' and ``Sarah's preferred
airport is SEA.'' A question about the user's boss and flights may share words
with the first statement but not with the answer statement.

The derived graph connects these records through entities and typed
relations. Retrieval starts with a small number of lexical seeds, then spreads
through a bounded frontier for a limited depth. The default design uses no
learned graph weights; trust, relation priors, recency, and hop decay are
versioned deterministic inputs.

The useful output is not only ``SEA.'' It is the path that explains why the
record was nominated: the query reached the boss relation, that relation
identified Sarah, and Sarah's accepted airport relation led to SEA. Every
edge cites its source record.

Graph candidates and direct lexical candidates are combined with a
rank-fusion method rather than pretending their scores have the same meaning.
The graph remains optional, and direct record recall remains available when
an item cannot be represented as an edge.

In the paper's specific 10,002-record probe, direct top-10 recall missed the
older airport record, while graph recall returned it at rank 2 after visiting
two edges. This validates the constructed mechanism, not general question
answering quality.

\boundary{Path explanation proves how a result was reached, not that every
relation on the path is factually correct.}

\clearpage
\section{Recall Forensics: Explaining a Past Answer}
\articlelabel{11}
\coreidea{A memory system should be able to reconstruct why a specific result
appeared at a specific time.}

Traditional retrieval evaluation asks whether the system found something
relevant. Recall forensics asks a deployment question: if a surprising memory
appeared last Tuesday, what durable evidence explains its ranking, source,
lifecycle state, and integrity?

The \system{} paper defines five checks. Attribution recomputes ranking inputs
and ordering. Provenance follows the result back to admission, episode, and
origin. Lifecycle replay checks whether the item was active at that turn.
Decision replay reconstructs the logged selection. Integrity testing mutates
selected evidence fields to see whether verification notices.

In the content-addressed evaluation snapshot, an independent standard-library
auditor recomputed 14 of 14 ranking decisions and reconstructed all 68 returned
result pairs. It detected all 75 selected single-row mutations. Eleven stable
turns were re-executed exactly.

Three earlier turns were deliberately excluded from exact engine
re-execution because a later correction or deletion changed their canonical
inputs. That exclusion is part of the result: preserving deleted plaintext
only to make an audit demo look complete would violate the deletion contract.

Recall forensics is not the same as proving an answer true. It proves what the
measured evidence can reconstruct about why the system returned it.

\boundary{The evaluated workload is synthetic and bounded. Its mutation
coverage does not prove detection of every possible coherent database
rewrite.}

\clearpage
\section{A Hash Chain Is Useful, but It Needs an Anchor}
\articlelabel{12}
\coreidea{Tamper evidence stored beside the data cannot, by itself, prove that
the entire database was not replaced.}

\system{} appends important transitions to a per-subject SHA-256 hash chain.
Each event commits to the previous event hash. An edit, insertion, or interior
deletion breaks verification from the chain start.

This is valuable, but it has a limit. An attacker with complete database
control could truncate the tail or rewrite the entire chain and calculate new
hashes. If the verifier trusts only the modified file, there is no independent
reference for the old head.

The answer is an external checkpoint. A checkpoint records the event count,
sequence, and current hash, then moves that commitment to an independent
location such as append-only object storage or a timestamping service.

The paper's mutation experiment demonstrates the distinction directly. Tail
truncation was detected in zero of five trials using the chain alone and in
five of five when checked against an anchored head.

An incremental verification cache can make routine checks faster by verifying
only the new suffix. That cache is an optimization, not a trust anchor, because
an attacker who replaces the database may also replace a local cache.

\boundary{A hash chain provides tamper evidence relative to a trusted point.
It does not provide replication, availability after disk loss, trusted wall
clock time, or automatic recovery.}

\clearpage
\section{Confidence Is Not Trust}
\articlelabel{13}
\coreidea{One number should not decide whether information is relevant,
trusted, true, or authorized.}

AI systems often use the word ``confidence'' as if it described one property.
In practice, several different questions are being mixed together.

\system{} separates extraction confidence, source trust, record retrieval
score, graph activation, entity-merge confidence, and external-action outcome.
Each value has a limited job.

Extraction confidence records how strongly a deterministic parser recognized
a supported sentence form. Trust records where the statement came from and
whether it was reviewed. Retrieval scores order already eligible candidates.
Graph scores measure the strength of a typed path. Merge confidence describes
an identity proposal. Action outcomes report what was observed after an
external operation.

A high retrieval score cannot activate quarantined content. A high merge score
cannot merge two entities automatically. A confident memory cannot authorize
a file change. An external timeout is not converted into ``60 percent
successful''; it becomes uncertain because the outside state is unknown.

This separation makes the system less convenient to summarize with one
dashboard number, but much easier to reason about. Every value answers a
specific question and has a defined authority boundary.

\boundary{The current ranking and merge values are deterministic heuristics,
not calibrated probabilities of truth or correctness.}

\clearpage
\section{Ambiguity Should Stay Visible}
\articlelabel{14}
\coreidea{When the system cannot justify a conclusion, abstention or review is
better than a hidden guess.}

Language is ambiguous. Identities are ambiguous. External systems can fail in
ambiguous ways. A memory architecture becomes fragile when it compresses all
of these situations into a confident-looking answer.

The current \system{} extractor is intentionally narrow. Unsupported phrasing
creates a source episode but may create no semantic record. Questions are not
treated as assertions. A caller can request a forced note, but that choice is
explicit and recorded.

Unstructured conflicts may remain as multiple records rather than being
silently merged. Entity-resolution candidates wait for a reviewer. Exact name
or alias overlap can support a proposal, but it does not prove two entities
are the same.

External ambiguity receives its own states. If an operation fails after
dispatch, generic code may not know whether the provider completed it. The
system records uncertainty or recovery required instead of retrying blindly
and risking a duplicate effect.

This approach gives users more visible states to understand. That is a design
cost, but it reflects reality. A single green check mark is not useful when
the system cannot prove what happened.

\boundary{Abstention reduces silent corruption but can reduce convenience and
recall coverage. The product must make review workflows understandable rather
than leaving uncertainty as an unexplained error.}

\clearpage
\section{Information Is Not Authority}
\articlelabel{15}
\coreidea{Remembering a fact should not grant permission to act on it.}

An assistant may remember that a report belongs in a certain folder. That
memory can help prepare a proposal. It should not, by itself, authorize the
assistant to overwrite a file.

\system{} separates memory from external authority. Recalled records can
inform an action plan, but enforced write operations require a trusted,
host-attested authority reference. Untrusted records may contribute context
without becoming permission.

This distinction matters for prompt injection. A web page might tell an agent
to edit a configuration file. Even if the content is stored for review, it
cannot manufacture the user authority required by the guarded action engine.

The agent-facing interface can stage tools and inspect state, but it cannot
label its own proposal as a user instruction or create approval material. The
reviewer-side interface creates authority from the user's actual task.

This is one of the main reasons \system{} is more than a retrieval component.
It connects memory evidence to an effect-control plane while keeping the two
roles distinct.

\boundary{The guarantee applies only when model-requested effects are routed
through the registered broker. Direct shell, filesystem, database, or provider
access can bypass the gate and must be restricted by deployment policy.}

\clearpage
\section{Approve the Exact Plan, Not the General Intention}
\articlelabel{16}
\coreidea{Approval should bind to the precise operation that was reviewed.}

``Update my report'' is an intention, not a complete execution plan. The file
may change after review. An adapter may be upgraded. A path may resolve
differently. The proposed content may no longer match what the user saw.

\system{} prepares a canonical plan containing the operation, arguments,
preview, preconditions, adapter manifest, dependencies, evidence references,
and policy version. It calculates a digest over that complete plan.

Reviewer approval uses HMAC-SHA256 to bind the transaction identifier, plan
digest, approver label, issue time, expiry time, and a random nonce. The
approval applies to one exact plan for a limited period, not to future
variations of the same idea.

Before execution, the engine recomputes the plan and approval commitments,
checks the audit chain, confirms the adapter did not change, and revalidates
preconditions. Each operation is checked again immediately before its effect.

For the reference filesystem adapter, a changed file digest causes the plan to
stop instead of stretching the old approval to cover new state. After a write,
the adapter independently reads the result and verifies its expected digest.

\boundary{An HMAC proves possession of a shared secret, not a person's legal
identity. Multi-user or regulated deployment would require stronger identity,
revocation, key custody, and policy administration.}

\clearpage
\section{``Uncertain'' Is a Responsible Result}
\articlelabel{17}
\coreidea{When an external call may have succeeded, pretending failure and
retrying can be more dangerous than admitting uncertainty.}

Database operations can often commit or roll back atomically. External calls
cannot. A provider may receive a request, perform the effect, and disconnect
before returning a response. The caller sees an error but does not know the
outside state.

\system{} persists execution state before dispatch and does not hold a SQLite
transaction open across the external call. If the call raises after dispatch,
the action becomes uncertain. If the process restarts during a committing or
compensating operation, the transaction becomes recovery required.

Provider-specific idempotency lookup or authoritative state inspection may
later resolve the outcome. Generic code does not retry blindly, because a
second request could duplicate the first effect.

For multi-step plans, previously verified effects may be compensated in
reverse order. Compensation runs only when current state still matches the
engine's verified after-state. If another actor changed the resource, the
engine refuses to overwrite that later work.

The final result may be committed, compensated, partial, uncertain, or
recovery required. These states are less tidy than success and failure, but
they preserve what the system actually knows.

\boundary{A generic control plane cannot resolve ambiguity when a provider
offers neither idempotency records nor an authoritative way to inspect the
resulting state.}

\clearpage
\section{What Aetnamem Encryption Protects Today}
\articlelabel{18}
\coreidea{Encryption claims are meaningful only when they name the platform,
files, process state, and remaining plaintext.}

The current \system{} release does not use SQLCipher and does not provide one
uniform live-database encryption design across macOS, Windows, and Linux.

On macOS, the desktop launcher can seal an idle database using OpenSSL
AES-256-CBC with PBKDF2 and an HMAC-SHA256 sidecar. The secret is held in the
login Keychain. Before opening the database, the HMAC is checked; during use,
SQLite operates on a plaintext database in a temporary directory.

The launcher checkpoints and reseals after mutations, subject to a short rate
limit, and force-seals during a clean shutdown. This provides useful
confidentiality when an attacker obtains only the idle sealed files without
the Keychain secret.

It is not transparent database encryption. The live database and WAL are
plaintext while the service runs. The ciphertext and HMAC sidecar are not
published atomically. The same secret currently serves two cryptographic
roles, and archive or workspace files need separate protection.

Windows and Linux currently store ordinary SQLite files. Their at-rest
confidentiality depends on operating-system controls such as BitLocker or
LUKS when those controls are enabled and managed correctly.

\boundary{Encryption protects data against specific storage threats. It does
not protect a database from every process running as the unlocked user, prove
memory correctness, or replace tested backup and key-recovery procedures.}

\clearpage
\section{Scale Claims Should Follow Evidence}
\articlelabel{19}
\coreidea{An architecture can be designed for growth without pretending that
unmeasured capacity has already been proven.}

\system{} uses three mechanisms to avoid obvious unbounded work in hot recall:
candidate-capped direct ranking, bounded graph traversal, and optional cold
partitions for inactive derived edges. Incremental audit verification avoids
rehashing an already validated prefix on every routine check.

These choices make larger memory stores plausible. They do not establish a
supported maximum. SQLite file size, full-text selectivity, cache pressure,
audit growth, backup duration, archive churn, concurrent writes, and hardware
all affect real deployments.

The paper therefore reports what it measured. In one Apple M2 probe with
10,002 records, direct recall had a median of 17.139 milliseconds and graph
recall 24.799 milliseconds. Graph recall found the constructed two-hop target
at rank 2; direct top-10 recall did not. The candidate cap remained 200.

That experiment supports the intended bounded-recall and multi-hop mechanism
for one synthetic scenario. It is not a million-record benchmark, a
multilingual relevance evaluation, a sustained concurrency test, or evidence
of state-of-the-art answer quality.

Engineering credibility improves when design projections and validated
results remain separate. The next useful evidence is not a larger adjective;
it is a broader workload, reproducible measurements, and published failure
conditions.

\boundary{Performance numbers vary by host and run. The more reproducible
claims are the configured bounds, returned rank, graph path, and visited-edge
count.}

\clearpage
\section{The End Goal: Governed Memory Infrastructure}
\articlelabel{20}
\coreidea{The long-term value is a memory layer that remains useful as models,
interfaces, and applications change.}

\system{} is designed so that memory does not belong to one language model.
The model can be local or remote and can change over time. Canonical records,
trust transitions, graph lineage, audit evidence, and approvals remain outside
the model.

This makes the system potentially useful in several roles: personal assistant
memory, memory for agents using MCP or an SDK, governed intake from external
channels, and decision memory around engineering workflows. The application
may use recalled records to augment a prompt, but retrieval augmentation is
only one integration point.

The more precise description is a \textbf{governed memory and effect-control
plane}. It keeps accepted information durable, separates suspicious material,
shows why a result surfaced, distinguishes information from authority, and
records uncertainty when an external outcome cannot be proven.

The architecture does not depend on claims that every component is new. Full
text search, graphs, provenance, hash chains, approvals, and compensating
actions are established techniques. The systems contribution is their
composition around inspectable memory transitions and approval-bound effects.

The remaining work is concrete: stronger cross-platform encryption, broader
retrieval-quality evaluation, larger scale tests, operating-system sandboxing,
better adapters, and independent security review. Those improvements matter
more than presenting the system as finished.

The intended user experience is simple: remember what was accepted, show
where it came from, keep unreviewed content separate, ask before changing the
world, and admit when the evidence is incomplete.

\boundary{\system{} is an evolving open-source system. Its potential should
be judged against implemented behavior, reproducible evidence, and explicit
limitations rather than unsupported claims of universal novelty.}

\end{document}
